---
title: "Non-proportional hazards and non-linear exposure-response functions"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
  word_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(targets)
library(knitr)
```


# Introduction
## Background
In this case study, we employed a Cox proportional hazards model to investigate the relationship between an exposure and health outcomes. During the model diagnostics, we identified non-proportional hazards for the exposure of interest, indicating that the effect estimate on the hazard rate varied over time.

To address this violation of the proportional hazards assumption, we introduced time-varying effects through step functions. This was implemented using an interaction term with time strata, allowing us to account for the changing impact of air pollution across different time groups.

Our primary aim was to assess the non-linearity of the exposure-response relationship over the range of air pollution values. To achieve this, we modeled air pollution in quintiles and included an interaction term for the age strata. 

## Methods and Data
We use the veterans cancer data from the `timedep` vignette from the `survival` package by Terry Therneau. We also follow suggestions from Frank Harrell author of the Regression Modelling Strategies book and `rms` package.

```{r}
library(survival)
library(rms)
```


### Statistical Analysis
#### 1. Fitting the Cox Model
```{r, veteran1}
vfit <- coxph(Surv(time, status) ~ trt + prior + karno, veteran)
summary(vfit)
```

#### 2. Checking the Proportional Hazards Assumption
- the proportional hazard assumption is precisely that β(t) = β, i.e., the coeffcient does not change over time.

```{r, veteran1b}
zp <- cox.zph(vfit, transform = function(time) log(time + 20))
zp
plot(zp[3], resid = FALSE)
abline(0, 0, lty = 3)
abline(h = vfit$coef[3], lwd = 2, lty = 3)
```

- Solid line = the effect estimate for Karnofsy score, versus time, in the Veteran's cancer data set, as estimated by cox.zph, along with confidence intervals. Dotted lines are the overall estimate from a Cox model with karno as a time-fixed effect, and a reference line at 0.
- Early on it has a large negative effect, but by 200 days this has waned and is not much different from zero.

#### 3. Addressing Non-Proportional Hazards
- Implementation of time-varying effects
- Age stratification approach
- One of the simplest extensions is a step function for β(t), i.e., different coeffcients over different time intervals.


```{r, split}
vet2 <- survSplit(Surv(time, status) ~ ., data= veteran, cut=c(90, 180), 
                  episode= "tgroup", id="id")
vet2[1:7, c("id", "tstart", "time", "status", "tgroup", "age", "karno")]
vfit2 <- coxph(Surv(tstart, time, status) ~ trt + prior +
                 karno:strata(tgroup), data=vet2)
vfit2
cox.zph(vfit2)
# ADD THE NEW COEFFS to the plot 
plot(zp[3], resid=FALSE)    # a plot for the 3rd variable in the fit
abline(0,0, lty=3)
# add the effect estimate for the age stratum 0-90
abline(h= vfit2$coef[3], lwd=2, lty=1, col = 'green')
# add the effect estimate for the age stratum 90-180
abline(h= vfit2$coef[4], lwd=2, lty=1, col = 'blue')
# add the effect estimate for the age stratum 180+
abline(h= vfit2$coef[5], lwd=2, lty=1, col = 'red')
```

- This plot is showing the different HRs estimated for the different age strata. There is a strong negative coefficient for the time period 0-90, then a non-significant positive coefficient for 90-180 and subsequently a non-significant slight negative.
- An alternate way to plot this is shown below, similar to the termplot function

```{r, like-termplot}
x <- 0:max(veteran$karno)
y <- x * vfit2$coef[3]
plot(x,y, ylim = c(-4,2), xlim = c(0,100), xlab = 'Karno', ylab = 'log HR')
y2 <- x * vfit2$coef[4]
points(x,y2, col = 'green')
y3 <- x * vfit2$coef[5]
points(x,y3, col = 'blue')
```

#### 4. Model Refinement
- Non-linearity using a restricted cubic spline

```{r, non-linear}
# if the exposure of interest may be non-linear, 
vfit2.11 <- coxph(Surv(time, status) ~ trt + prior + rcs(karno, df = 4), data=vet2)
vfit2.11

termplot(vfit, terms = 3, se = T, ylim = c(-5,2))
termplot(vfit2.11, terms = 3, se = T, ylim = c(-5,2))
```

- Interaction term between time and non-linear effect estimate

```{r, non-linear time varying}
library(ggplot2)
# if the exposure of interest may be non-linear, 
vfit2.12 <- cph(Surv(time, status) ~ trt + prior + rcs(karno, df = 3) + strat(tgroup), data=vet2)
dd <- datadist(vet2)
options(datadist = 'dd')
par(mfrow = c(2,2))
p <- Predict(vfit2.12,  karno, tgroup = 1)
ggplot(p)
p <- Predict(vfit2.12,  karno, tgroup = 2)
ggplot(p)
p <- Predict(vfit2.12,  karno, tgroup = 3)
ggplot(p)


# trt = seq(min(vet2$trt), max(vet2$trt), length.out=1),
             # prior =  seq(min(vet2$prior), max(vet2$prior), length.out=2),
             # karno = seq(min(vet2$karno), max(vet2$karno), length.out=10),
             # tgroup=1
             # )

# p2 <- data.frame(p)
# with(p2, plot(karno, yhat))

```

```{r, non-linear with manual df}
vfit2.12 <- coxph(Surv(time, status) ~ trt + prior + rcs(karno, df = 5):strata(tgroup), data=vet2)

# Create prediction data
karno_seq <- seq(min(vet2$karno), max(vet2$karno), length.out = 100)
pred_data <- data.frame(
  karno = rep(karno_seq, 3),
  tgroup = rep(1:3, each = 100),
  trt = 0,  
  prior = mean(vet2$prior),
  tstart = rep(c(0, 90, 180), each = 100),
  time = rep(c(90, 180, 411), each = 100)
)

# Get predictions
preds <- predict(vfit2.12, newdata = pred_data, se.fit = TRUE)

# Set up multi-panel plot
par(mfrow = c(2,2))

# Plot for each time group
for(tg in 1:3) {
  subset_data <- data.frame(
    karno = karno_seq,
    loghr = preds$fit[pred_data$tgroup == tg],
    lower = preds$fit[pred_data$tgroup == tg] - 1.96 * preds$se.fit[pred_data$tgroup == tg],
    upper = preds$fit[pred_data$tgroup == tg] + 1.96 * preds$se.fit[pred_data$tgroup == tg]
  )
  
  plot(subset_data$karno, subset_data$loghr, type = "l",
       ylim = c(-5, 2),  # matching your original scale
       xlab = "Karnofsky Score",
       ylab = "log(HR)",
       main = paste("Time Period", tg, "\n", 
                   ifelse(tg == 1, "0-90 days",
                          ifelse(tg == 2, "90-180 days", ">180 days"))),
       lwd = 2)
  
  # Add confidence intervals
  lines(subset_data$karno, subset_data$lower, lty = 2)
  lines(subset_data$karno, subset_data$upper, lty = 2)
  
  # Add reference line at log(HR) = 0
  abline(h = 0, col = "gray", lty = 3)
}

# Add overall title in fourth panel
plot.new()
text(0.5, 0.5, "Non-linear effect of\nKarnofsky Score\nby time period", 
     cex = 1.2, font = 2)


```

- non-linearity can also be estimated using ordinal factor variable based on quantiles

```{r as function}
# Function to create prediction data and plot for a stratified model with splines
plot_stratified_spline_effect <- function(
  model,              # Fitted coxph model
  data,               # Original dataset
  spline_var,         # Name of the spline variable (e.g., "karno")
  strata_var,         # Name of the stratification variable (e.g., "tgroup")
  other_vars = NULL,  # List of other variables in model and their reference values
  n_points = 100,     # Number of points for prediction
  ylim = c(-5, 2),    # Y-axis limits
  plot_title = "Non-linear effect across strata",
  strata_labels = NULL # Optional custom labels for strata
) {
  # Extract variable from string name
  spline_var_values <- data[[spline_var]]
  strata_var_values <- data[[strata_var]]
  
  # Get strata levels
  strata_levels <- sort(unique(strata_var_values))
  n_strata <- length(strata_levels)
  
  # Create sequence for spline variable
  var_seq <- seq(min(spline_var_values, na.rm = TRUE), 
                max(spline_var_values, na.rm = TRUE), 
                length.out = n_points)
  
  # Set up reference values for other variables
  ref_values <- list()
  if (!is.null(other_vars)) {
    for (var_name in names(other_vars)) {
      if (is.null(other_vars[[var_name]])) {
        # Use mean or most common value if not specified
        if (is.numeric(data[[var_name]])) {
          ref_values[[var_name]] <- mean(data[[var_name]], na.rm = TRUE)
        } else {
          tab <- table(data[[var_name]])
          ref_values[[var_name]] <- names(tab)[which.max(tab)]
        }
      } else {
        ref_values[[var_name]] <- other_vars[[var_name]]
      }
    }
  }
  
  # Create prediction data frame
  pred_data <- data.frame(
    matrix(nrow = n_points * n_strata, ncol = length(ref_values) + 2)
  )
  
  # Set column names
  colnames(pred_data) <- c(spline_var, strata_var, names(ref_values))
  
  # Fill prediction data
  pred_data[[spline_var]] <- rep(var_seq, n_strata)
  pred_data[[strata_var]] <- rep(strata_levels, each = n_points)
  
  # Add reference values for other variables
  for (var_name in names(ref_values)) {
    pred_data[[var_name]] <- rep(ref_values[[var_name]], n_points * n_strata)
  }
  
  # Get predictions
  preds <- predict(model, newdata = pred_data, se.fit = TRUE)
  
  # Create plots
  if (is.null(strata_labels)) {
    strata_labels <- paste("Stratum", strata_levels)
  }
  
  # Set up multi-panel plot
  panel_rows <- ceiling(sqrt(n_strata + 1))
  panel_cols <- ceiling((n_strata + 1) / panel_rows)
  par(mfrow = c(panel_rows, panel_cols))
  
  # Plot for each stratum
  for (i in 1:n_strata) {
    idx <- pred_data[[strata_var]] == strata_levels[i]
    subset_data <- data.frame(
      x = var_seq,
      y = preds$fit[idx],
      lower = preds$fit[idx] - 1.96 * preds$se.fit[idx],
      upper = preds$fit[idx] + 1.96 * preds$se.fit[idx]
    )
    
    plot(subset_data$x, subset_data$y, type = "l",
         ylim = ylim,
         xlab = spline_var,
         ylab = "log(HR)",
         main = strata_labels[i],
         lwd = 2)
    
    # Add confidence intervals
    lines(subset_data$x, subset_data$lower, lty = 2)
    lines(subset_data$x, subset_data$upper, lty = 2)
    
    # Add reference line at log(HR) = 0
    abline(h = 0, col = "gray", lty = 3)
  }
  
  # Add overall title in last panel
  plot.new()
  text(0.5, 0.5, plot_title, cex = 1.2, font = 2)
  
  # Return the prediction data for further use if needed
  invisible(list(
    pred_data = pred_data,
    predictions = preds
  ))
}

plot_stratified_spline_effect(
  model = vfit2.12,
  data = vet2,
  spline_var = "karno",
  strata_var = "tgroup",
  other_vars = list(
    trt = 0,
    prior = NULL  # This will use mean value automatically
  ),
  ylim = c(-5, 2),
  strata_labels = c("0-90 days", "90-180 days", ">180 days"),
  plot_title = "Non-linear effect of\nKarnofsky Score\nby time period"
)
```
## Results
### Initial Model Assessment
### Time-Varying Effects Analysis
### Final Model and Interpretation

## Discussion
### Key Findings
### Methodological Considerations
### Implications for Environmental Epidemiology

## References